\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,bm}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[numbers]{natbib}

\input{macros}

\title{The Astrophysical Structure Field Equation: Detailed Derivation, Term-by-Term Meaning, and Plain-Language Explanation}
\author{Fabio Facin}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This companion note explains the Astrophysical Structure Field (ASF) equation in
two parallel languages: a mathematical language suitable for a computational
astrophysics paper, and a plain-language version suitable for explaining the
idea to a non-specialist. The central equation is not presented as a new
fundamental law of gravity. It is a phenomenological subgrid/effective-potential
model: a way to test whether unresolved baryonic structure can be encoded as a
dimensionless field whose spatial gradients produce measurable corrections in
dwarf-galaxy simulations.
\end{abstract}

\section{The Core Idea in One Paragraph}

Gravity in a galaxy simulation is usually computed from mass density. The ASF
idea asks whether mass density alone misses something important: the way gas,
magnetic fields, turbulence, and temperature structure are organized below the
simulation's resolution. The model collects those unresolved features into a
dimensionless scalar field, $\SA$. It then tests whether spatial changes in
that field can create a small effective correction to the potential used by
particles in a simulation.

\paragraph{Plain-language version.}
Imagine two rooms with the same amount of furniture. If one room is tidy and
the other is full of barriers, hot spots, fans, and tangled wires, moving
through them will feel different even though the total amount of stuff is the
same. ASF asks a similar question for dwarf galaxies: can two galaxies with
similar mass distributions behave differently because their gas and internal
structure are organized differently?

\section{Starting Point: The Newtonian Potential}

The standard Newtonian gravitational potential, $\PhiN$, satisfies Poisson's
equation:
\begin{equation}
  \laplacian \PhiN = 4\pi G\rho.
  \label{eq:explainer_poisson}
\end{equation}
Here $\rho$ is mass density and $G$ is Newton's gravitational constant. The
operator $\laplacian$ measures how curved or locally bent the potential field
is. In ordinary Newtonian gravity, mass density is the source of that curvature.

\paragraph{Plain-language version.}
Mass tells gravity how to shape the potential. The potential is like a
landscape: particles roll downhill in it. Poisson's equation says that where
there is mass, the landscape bends.

\section{Why Add a Structure Field?}

Large galaxy simulations cannot resolve every cloud, shock, magnetic filament,
feedback bubble, turbulent eddy, or temperature gradient. These unresolved
features can still influence gas motion and halo response through pressure
support, turbulence, and feedback. Modern galaxy simulations already use
subgrid prescriptions for unresolved physics \citep{Pillepich2018IllustrisTNG}.
ASF follows that modeling philosophy, but focuses on whether unresolved
structure can be represented as an effective-potential correction.

The proposed ASF-modified equation is
\begin{equation}
  \laplacian \Phieff = 4\pi G\rho + \lambdaA \laplacian \SA.
  \label{eq:explainer_asf_poisson}
\end{equation}
The normal mass-density term remains unchanged. The new term is
$\lambdaA\laplacian\SA$, where $\SA$ is dimensionless and $\lambdaA$ has units
of potential, or squared velocity.

\paragraph{Plain-language version.}
The model does not say gravity is wrong. It says: keep normal gravity, but add
a controlled correction that represents unresolved structure. Then test whether
that correction matters.

\section{From the Poisson Form to the Effective Potential}

Subtract Equation~\ref{eq:explainer_poisson} from
Equation~\ref{eq:explainer_asf_poisson}:
\begin{equation}
  \laplacian(\Phieff - \PhiN) = \lambdaA \laplacian \SA.
\end{equation}
If the two potentials are compared under compatible boundary conditions, this
implies the approximate effective-potential form
\begin{equation}
  \Phieff \approx \PhiN + \lambdaA \SA.
  \label{eq:explainer_phi_eff}
\end{equation}
The word ``approximate'' matters. Poisson equations are sensitive to boundary
conditions. The compact expression is a useful implementation form, but a
serious paper must state which boundary conditions are used.

\paragraph{Plain-language version.}
The correction says: start with the normal gravity landscape, then add a small
extra landscape shaped by the structure field. Where the structure field is
flat, nothing changes. Where it changes sharply, particles feel an extra push.

\section{The Acceleration Law}

Particles respond to gradients of the potential:
\begin{equation}
  \bm{a}_{\mathrm{eff}} = -\grad\Phieff.
\end{equation}
Using Equation~\ref{eq:explainer_phi_eff},
\begin{equation}
  \bm{a}_{\mathrm{eff}}
  = -\grad\PhiN - \lambdaA \grad\SA.
  \label{eq:explainer_acceleration}
\end{equation}
This is one of the most important forms of the model. The ASF contribution is
not controlled by the absolute value of $\SA$, but by its gradient. A uniform
structure field produces no extra acceleration.

\paragraph{Plain-language version.}
A perfectly flat road does not make a ball roll. A sloped road does. ASF works
the same way: only changes in the structure field matter.

\section{The Structure Field Definition}

The v1 ASF field is
\begin{equation}
  \SA =
  \ln\left(\frac{\rho}{\rho_0}\right)
  + \chi_B \frac{B^2}{2\mu_0 P_{\mathrm{gas}}}
  + \chi_\omega \frac{|\grad \times \bm{v}|^2}{\Omega_0^2}
  + \chi_T \frac{L_0^2 |\grad T|^2}{T_0^2}.
  \label{eq:explainer_structure}
\end{equation}
This is a modeling ansatz, not a first-principles theorem. Each term is chosen
because it represents a physically meaningful type of unresolved organization:
density contrast, magnetic support, vortical/turbulent structure, and thermal
structure.

\section{Term 1: Density Contrast}

The density term is
\begin{equation}
  \SA^{(\rho)} = \ln\left(\frac{\rho}{\rho_0}\right).
\end{equation}
The ratio $\rho/\rho_0$ compares the local density to a reference density. The
logarithm makes the term dimensionless and compresses large density contrasts.
This is useful because astrophysical densities can vary by many orders of
magnitude.

\paragraph{Where it comes from.}
Many physical models use logarithmic density contrast because it measures
relative structure rather than absolute density. In the ASF model, this term
asks whether density organization beyond the smooth mass profile contributes to
the effective correction.

\paragraph{Plain-language version.}
This term asks: is this region unusually dense or unusually empty compared with
the typical background? The logarithm keeps extreme values from dominating too
easily.

\section{Term 2: Magnetic Support}

The magnetic term is
\begin{equation}
  \SA^{(B)} = \chi_B \frac{B^2}{2\mu_0 P_{\mathrm{gas}}}.
\end{equation}
The quantity $B^2/(2\mu_0)$ is magnetic pressure. Dividing by gas pressure,
$P_{\mathrm{gas}}$, creates a dimensionless ratio comparing magnetic support
to ordinary thermal pressure.

\paragraph{Where it comes from.}
In magnetohydrodynamics, magnetic fields contribute stresses and pressure-like
support. The ASF model does not claim magnetic fields create gravity. Instead,
it asks whether unresolved magnetic support can be represented as part of an
effective structure correction when the simulation does not resolve those
stresses directly.

\paragraph{Plain-language version.}
Gas pressure is like ordinary pushing from hot gas. Magnetic pressure is like
extra stiffness from magnetic fields. This term asks whether magnetic stiffness
is important compared with normal gas pressure.

\section{Term 3: Vorticity and Turbulent Organization}

The vortical term is
\begin{equation}
  \SA^{(\omega)} =
  \chi_\omega \frac{|\grad \times \bm{v}|^2}{\Omega_0^2}.
\end{equation}
The curl of the velocity field, $\grad\times\bm{v}$, measures local rotation or
swirl. Squaring its magnitude gives a positive scalar measure of vortical
structure. Dividing by $\Omega_0^2$ makes the term dimensionless.

\paragraph{Where it comes from.}
Turbulence and feedback-driven motion are central uncertainties in dwarf-galaxy
simulations. Vorticity is not the whole story of turbulence, but it is a
compact scalar diagnostic for organized rotational motion. ASF uses it as a
first testable proxy.

\paragraph{Plain-language version.}
This term measures how much the gas is swirling. A calm pond and a whirlpool
can contain the same amount of water, but motion through them is different.

\section{Term 4: Thermal Gradients}

The thermal term is
\begin{equation}
  \SA^{(T)} =
  \chi_T \frac{L_0^2 |\grad T|^2}{T_0^2}.
\end{equation}
The gradient $\grad T$ measures how rapidly temperature changes across space.
Multiplying by $L_0^2$ and dividing by $T_0^2$ makes the term dimensionless.

\paragraph{Where it comes from.}
Strong temperature gradients can signal shocks, feedback bubbles, cooling
fronts, or multiphase gas boundaries. These are precisely the types of
structure that may be unresolved in a coarse simulation.

\paragraph{Plain-language version.}
This term asks whether the galaxy has sharp hot-cold boundaries. Sharp
temperature changes can mark energetic feedback or disturbed gas.

\section{What the Coefficients Mean}

The parameters $\chi_B$, $\chi_\omega$, and $\chi_T$ are dimensionless weights.
They do not have the same status as physical constants. They are modeling
parameters that must be scanned, constrained, and possibly rejected.

The coupling $\lambdaA$ controls the overall size of the potential correction:
\begin{equation}
  \lambdaA = \epsilon_A v_c^2.
\end{equation}
Here $v_c$ is a characteristic circular speed and $\epsilon_A$ is dimensionless.
For a conservative first model, $\epsilon_A$ should usually be small unless
there is strong evidence for a larger correction.

\paragraph{Plain-language version.}
The $\chi$ values decide which kind of structure matters most. The $\lambdaA$
value decides how strongly structure affects motion at all. If these numbers
can be chosen freely to fit anything, the model is weak. If only a narrow,
physically reasonable range works, the model becomes more serious.

\section{Effective-Density Interpretation}

Equation~\ref{eq:explainer_asf_poisson} can be rewritten as
\begin{equation}
  \laplacian\Phieff = 4\pi G(\rho + \rhoStruct),
\end{equation}
where
\begin{equation}
  \rhoStruct = \frac{\lambdaA}{4\pi G}\laplacian\SA.
  \label{eq:explainer_rho_struct}
\end{equation}
This is a useful diagnostic, but it must be interpreted carefully.
$\rhoStruct$ is not literal new matter. It is the effective source that would
produce the same potential correction in a Poisson equation.

\paragraph{Plain-language version.}
This is like saying, ``If I wanted to imitate the effect of structure using an
extra fake density map, what fake density would I need?'' It is a diagnostic
tool, not a claim that extra mass really exists.

\section{Why This Is a Computational Astrophysics Model}

The model belongs most naturally to computational astrophysics because its
validity depends on simulations and diagnostics, not on elegance alone. The
core tests are:
\begin{enumerate}
  \item Does $\lambdaA=0$ recover the Newtonian-only simulation?
  \item Does uniform $\SA$ produce zero extra acceleration?
  \item Does a structured $\SA$ produce measurable but controlled changes?
  \item Are those changes stable under resolution, smoothing, and boundary
  condition tests?
  \item Does the correction improve any diagnostic without destroying others?
\end{enumerate}

\section{What Would Count as Evidence}

ASF becomes more credible if the following are true:
\begin{itemize}
  \item small corrections alter central particle orbits without ruining outer
  halo behavior;
  \item similar mass maps but different structure maps produce measurably
  different rotation-curve diagnostics;
  \item useful behavior occurs only in a constrained parameter range;
  \item the correction correlates with physical structure rather than arbitrary
  curve fitting;
  \item failure regions are reported honestly.
\end{itemize}

\section{What Would Count Against the Model}

ASF should be rejected or revised if:
\begin{itemize}
  \item it requires unphysically large $\lambdaA$;
  \item it is numerically unstable for reasonable fields;
  \item it double-counts effects already resolved by hydrodynamic or MHD
  equations;
  \item results are dominated by boundary conditions or smoothing choices;
  \item it can fit anything by changing too many parameters;
  \item it worsens rotation curves, orbit behavior, or conservation diagnostics.
\end{itemize}

\section{The Most Honest Derivation Statement}

The current derivation is a modeling derivation:
\begin{enumerate}
  \item begin with the accepted Newtonian Poisson equation;
  \item introduce an effective potential for unresolved structure;
  \item require the correction to vanish in smooth or disabled limits;
  \item require the new field to be dimensionless;
  \item choose physically motivated scalar diagnostics for density contrast,
  magnetic support, vorticity, and thermal gradients;
  \item test whether the resulting correction is stable, constrained, and
  useful.
\end{enumerate}
This is not yet a first-principles derivation from the Einstein equations,
full MHD, or a turbulence closure theory. A future stronger version should try
to derive $\SA$ from coarse-grained stress terms, Jeans-equation corrections, or
an effective action.

\section{A Normal-Person Explanation}

The ASF equation is a way to ask whether the hidden organization inside a dwarf
galaxy matters. Normal gravity cares about mass. But real dwarf galaxies also
contain gas, magnetic fields, hot and cold regions, and turbulent motion. A
computer simulation cannot see every small detail. ASF makes a map of those
hidden details and asks whether changes in that map cause small changes in how
particles move.

The idea is not, ``I discovered new gravity.'' The idea is, ``Maybe unresolved
structure leaves a measurable fingerprint in the effective potential used by a
simulation.'' The research project is to test that fingerprint carefully.

\section{Glossary}

\begin{center}
\begin{tabular}{ll}
\toprule
Symbol & Meaning \\
\midrule
$\PhiN$ & Standard Newtonian gravitational potential \\
$\Phieff$ & Effective potential including ASF correction \\
$\rho$ & Mass density \\
$\SA$ & Dimensionless Astrophysical Structure Field \\
$\lambdaA$ & Coupling strength with units of squared velocity \\
$B$ & Magnetic field magnitude \\
$P_{\mathrm{gas}}$ & Gas pressure \\
$\grad\times\bm{v}$ & Vorticity, or local swirling motion \\
$T$ & Gas temperature \\
$\chi_B,\chi_\omega,\chi_T$ & Dimensionless weights for structure terms \\
$\rhoStruct$ & Effective source-density diagnostic, not literal matter \\
\bottomrule
\end{tabular}
\end{center}

\section{Conclusion}

ASF is best understood as a disciplined computational question: can unresolved
baryonic structure be compressed into a dimensionless scalar field that
produces small, testable effective-potential corrections? The equation is
plausible enough to simulate, but it earns scientific value only through
derivation refinement, parameter discipline, stability tests, and honest
failure criteria.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
