\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,bm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[numbers]{natbib}

\input{macros}

\title{A Phenomenological Astrophysical Structure Field Model for Dwarf-Galaxy Halo Diversity}
\author{Fabio Facin}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We introduce the Astrophysical Structure Field (ASF), a phenomenological
subgrid/effective-potential model designed to test whether unresolved baryonic
organization can produce measurable corrections in dwarf-galaxy simulations.
The model preserves the Newtonian Poisson limit while adding a dimensionless
structure field, $\SA$, built from density contrast, magnetic support,
vorticity, and thermal gradients. The central claim is not that ASF is a new
law of gravity, but that unresolved structure may be represented as a compact
closure whose gradients alter effective accelerations. We describe the
mathematical form, dimensional consistency, controlled numerical experiments,
diagnostics, and failure criteria required to assess the model.
\end{abstract}

\section{Introduction}

Dwarf galaxies are sensitive laboratories for dark-matter halo structure
because their shallow potentials amplify the impact of baryonic feedback,
turbulence, gas outflows, magnetic fields, cosmic rays, and star-formation
history. The core-cusp and halo-diversity problems motivate models that can
separate mass-density effects from unresolved baryonic organization.

This paper develops ASF as a methods framework. The model is intentionally
phenomenological: it asks whether a dimensionless structure field can serve as
a controlled closure for unresolved organization in galaxy-scale simulations.

\section{Related Work}

Large cosmological simulations already use subgrid prescriptions for
unresolved galaxy-formation physics \citep{Pillepich2018IllustrisTNG}.
Baryonic feedback can alter central dark-matter profiles
\citep{BenitezLlambay2019EAGLECores}, and recent work continues to investigate
rotation-curve diversity, black holes, cosmic rays, and feedback in dwarf
systems \citep{Williams2025DARKexp,Gurvich2025FIRE,Thiele2025CosmicRaySubgrid}.
ASF is positioned within this computational closure tradition.

\section{Model Definition}

The Newtonian potential satisfies
\begin{equation}
  \laplacian \PhiN = 4\pi G\rho.
\end{equation}

ASF introduces an effective potential
\begin{equation}
  \laplacian \Phieff = 4\pi G\rho + \lambdaA \laplacian \SA,
  \label{eq:asf_poisson}
\end{equation}
which implies, under compatible boundary conditions,
\begin{equation}
  \Phieff \approx \PhiN + \lambdaA \SA.
  \label{eq:effective_potential}
\end{equation}

The v1 structure field is
\begin{equation}
  \SA =
  \ln\left(\frac{\rho}{\rho_0}\right)
  + \chi_B \frac{B^2}{2\mu_0 P_{\mathrm{gas}}}
  + \chi_\omega \frac{|\grad \times \bm{v}|^2}{\Omega_0^2}
  + \chi_T \frac{L_0^2 |\grad T|^2}{T_0^2}.
  \label{eq:structure_field}
\end{equation}
Each term is dimensionless. The coefficients $\chis$ control the relative
importance of magnetic, vortical, and thermal structure.

\section{Dimensional Analysis and Limits}

Since $\SA$ is dimensionless, $\lambdaA$ must have units of squared velocity:
\begin{equation}
  [\lambdaA] = L^2T^{-2}, \qquad \lambdaA = \epsilon_A v_c^2.
\end{equation}
The acceleration is
\begin{equation}
  \bm{a}_{\mathrm{eff}} = -\grad \Phieff
  = -\grad \PhiN - \lambdaA \grad \SA.
  \label{eq:acceleration}
\end{equation}
If $\lambdaA=0$ or $\grad \SA = 0$, the correction vanishes. If $\SA$ is weak
or intentionally disabled, Equation~\ref{eq:asf_poisson} recovers the standard
Poisson equation.

The correction can also be written as an effective source density,
\begin{equation}
  \rhoStruct = \frac{\lambdaA}{4\pi G}\laplacian \SA,
\end{equation}
but $\rhoStruct$ is not literal matter. It is a closure representation of
unresolved structure.

\section{Numerical Method}

The first experiments use a two-dimensional grid, finite differences, and a
Jacobi or Gauss-Seidel Poisson solver. The numerical pipeline is:
\begin{enumerate}
  \item initialize $\rho$, $P_{\mathrm{gas}}$, $\bm{B}$, $\bm{v}$, and $T$;
  \item compute $\SA$ from Equation~\ref{eq:structure_field};
  \item solve $\laplacian \PhiN = 4\pi G\rho$;
  \item compute $\Phieff = \PhiN + \lambdaA \SA$;
  \item compute $\bm{a}_{\mathrm{eff}} = -\grad\Phieff$;
  \item evolve test particles or compute rotation-curve diagnostics.
\end{enumerate}

\section{Controlled Experiments}

The core experiment holds $\rho$ fixed while varying $\SA$. This tests whether
two systems with the same mass density but different unresolved organization
produce measurable differences in potential maps, acceleration maps, particle
orbits, and circular velocity diagnostics.

\section{Diagnostics}

For approximate axisymmetric diagnostics, the circular velocity profile is
\begin{equation}
  \vc(r) = \sqrt{r\frac{d\Phieff}{dr}}.
\end{equation}
Additional diagnostics include potential residuals, orbit divergence, energy
drift, central acceleration changes, and the parameter region in which the ASF
correction is small but measurable.

\section{Limitations and Failure Criteria}

ASF must be rejected or revised if it requires unphysically large $\lambdaA$,
is numerically unstable, worsens rotation-curve agreement, double-counts
resolved hydrodynamic/MHD effects, violates conservation behavior, or depends
strongly on arbitrary boundary conditions. The form of $\SA$ is not unique and
must eventually be derived from a coarse-graining, turbulence-closure, or
effective-theory argument.

\section{Conclusion}

ASF provides a structured way to test whether unresolved baryonic organization
can alter effective-potential diagnostics in dwarf-galaxy simulations. The
model is useful only if it remains constrained, falsifiable, and distinguishable
from existing baryonic feedback prescriptions.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
